Regression Cube Analysis of Cohort Study Data

### 2015-03-21 draft.tex
worked on structure of Related Work

\subsection{The Study of Health in Pomerania (SHIP)}
The \texttt{SHIP}, located in Northern Germany aims to aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%%
It does not focus on a specific disease, making the data set open for many diseases.
%%
Unique for the \texttt{SHIP} is the acquisition of medical image data per subject.
%%
A second cohort, \texttt{SHIP-TREND} was introduced in 2012.
%%
Data for both cohorts are examined in a in a 5-year time span.
%%
New parameters are added in each iteration, extending the range of investigated diseases.
%%
For the last acquisition (\texttt{SHIP-2} and \texttt{SHIP-TREND-0}), MRI scans are included into the cohort \cite{Hegenscheid2009, Ivanovska2014}.
%%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives were initiated.
%
%They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
%Data are collected without focus on a group of diseases.
%
%This allows to query the data regarding many diseases and conditions.
%
%
\section{Prior and Related Work}
\com{From VAST'14 Paper:}
\paragraph{Visual Analysis of Heterogenous Data.}
Zhang et al. \cite{Zhang2012} provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set.
%
Their definition of a cohort differs from the understanding of the term in an epidemiological context by denoting every parameter-divided subject group as individual cohort.
%
Due to the short paper length, detail is missing on the data types and their algorithms of identifying similar subjects or whether they employ statistical measures.
%
We employ the idea of adding variables via drag and drop into a canvas area.

@@

\section{Implementation} \label{implementation}
\begin{figure}[htb]
 \centering
 \includegraphics[width=2.8in]{figures/placeholder}
 \caption{
 \com{Describe the Bootstrap/Angular/D3 Frontend, Node backend, R/OCPU Backend.}
 }
  \label{fig:Implementation}
\end{figure}
%%
Relying on web technologies paid off in previous collaboration with our clinical exerts, since it allows for:
\begin{itemize}
	\item easy exchange of software, without the need of installing anything other than a modern web browser,
	\item fast iterative adaption of method changes, without the need of installing updates,
	\item outsourcing computation-heavy tasks to server-clusters, which do not have to be in local proximity.
\end{itemize}
%%
The ongoing transition of open-science software into the web spawned numerous projects, making state-of-the-art algorithms available in this domain.
%%
\paragraph{Front-End.}
%%
The front end is created using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%%
\texttt{Angular.js}\footnote{Open Source; Maintained by Google, \href{https://www.angularjs.org/}{\texttt{www.angularjs.org}}} abstracts web application into models and views, allowing for a responsive way to combine \texttt{HTML} and \texttt{Javascript}.
%%
\texttt{Angular.js} is forcing developers to write modularized code, which makes the components easier expandable while keeping the code maintainable by including unit tests.
%%
The page layout is handled using \texttt{Twitter Bootstrap}\footnote{Open Source; Maintained by Twitter, \href{http://getbootstrap.com}{\texttt{getbootstrap.com}}}, which also provides a rich set of UI elements with proper stylings.
%%
The 2D mosaic plot is implemented using \texttt{Data driven Documents}(\texttt{D3.js}), which is very popular in information visualization using vector graphics \cite{D3}.
%%
It provides fast and easy methods for binding data to graphical elements.
%%
The 3D plot is created using the WebGL-based \texttt{Threejs}\footnote{Open Source; Originally developed by R. Cabello, \href{http://threejs.org}{\texttt{threejs.org}}} library.
%%
We experimented with different ways for achieving the cube representation, including volume rendering, cube primitives for each data point and shader-based solutions.
%%
Proper open source volume rendering methods are not available for the web, the effort of writing one from scratch was not worth the counter value.
%%
Creating a cube primitive for each data point resulted in non-interactive frame-rates for data sets larger than 30 features (creating 30$^3$ cube primitives).
%%
We use a shader-based solution by rendering the cube as a sprite based particle system, allowing to customize color and opacity of every data point.
%%
It also is the fastest solution that we tested.

@@

\paragraph{Access and Source.}
%%
A running instance of the \emph{Regression Cube} prototype can be found under \href{http://regressioncube.herokuapp.com/}{\texttt{regressioncube.herokuapp.com}}.
%%
The source for the prototype is freely available at \texttt{Github}\footnote{\texttt{R}-based back-end: \href{https://github.com/paulklemm/regression-cube-r-package}{\texttt{github.com/paulklemm/regression-cube-r-package}}}\footnote{Front-End and Nodejs Webserver: \href{https://github.com/paulklemm/regression-cube-prototype}{\texttt{github.com/paulklemm/regression-cube-prototype}}}.
%%
Instructions and code on how to create a setup running the \emph{Regression Cube} statistical backend through a \texttt{Ubuntu} server using \texttt{OpenCPU} are referenced in the repository.
%%
The front-end can be deployed using \href{https://www.heroku.com/}{\texttt{www.heroku.com}} by cloning the repository into a Heroku homepage. 
%%
%The front-end source as well as the \texttt{Nodejs} backend can be found here: .
%%
% \com{Use VAST'14 as Guideline}
% \com{Cloud computing}
% \com{Volume Rendering vs Particles}
% \com{Speed is merely a matter of available server machines due to parallelization process.}
% \com{Security enabled by hashing files.}
% \com{Bootstrap/Angular/D3 Frontend, Node backend, R/OCPU Backend (Fig.~\ref{fig:Implementation})}

@@

\section{Summary and Conclusion}
\paragraph{Future Work.}
\com{Implementing of regression analysis on the graphics card.}
%%
% \begin{small}
	%\acknowledgments{Omitted due to blind review}
\acknowledgments{SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no. 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania. Whole-body MR imaging was supported by a joint grant from Siemens Healthcare, Erlangen, Germany and the Federal State of Mecklenburg-Vorpommern. The University of Greifswald is a member of the ‘Centre of Knowledge Interchange’ program of the Siemens AG. This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics. We thank Marko Rak and Klaus Toennies for providing the image detection data.}
% \end{small}
\clearpage
\newpage
\bibliographystyle{abbrv}
%%use following if all content of bibtex file should be shown
%\nocite{*}
\bibliography{bibliography}
\end{document}

>>>>

\subsection{The Study of Health in Pomerania (SHIP)}
The \texttt{SHIP}, located in Northern Germany aims to aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%%
It does not focus on a specific disease, making the data set open for many diseases.
%%
Unique for the \texttt{SHIP} is the acquisition of medical image data per subject.
%%
A second cohort, \texttt{SHIP-TREND} was introduced in 2012.
%%
Data for both cohorts are examined in a in a 5-year time span.
%%
New parameters are added in each iteration, extending the range of investigated diseases.
%%
For the last acquisition (\texttt{SHIP-2} and \texttt{SHIP-TREND-0}), MRI scans are included into the cohort \cite{Hegenscheid2009, Ivanovska2014}.
%%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives were initiated.
%
%They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
%Data are collected without focus on a group of diseases.
%
%This allows to query the data regarding many diseases and conditions.
%
%
\section{Prior and Related Work}
\com{Add section on most similar publications}
\com{TODO: Needs to be rewritten, this is only here for the references!}
\com{From VAST'14 Paper:}

\com{http://www.ii.uib.no/vis/publications/publication/2014/Angelelli14Interactive}
%\com{http://link.springer.com/chapter/10.1007/978-3-642-39146-0_1#page-1}
\com{Turkey Metrics Paper}

\paragraph{Data Mining in Epidemiology.}
\com{Problem: Only paper of co-authors in this section, will look one-sided!}
\com{Uli Niemann, Tommy Hielscher, Myra Spiliopoulou, Henry Völzke, and Jens-Peter Kühn. Can we classify the participants of a longitudinal epidemiological study from their previous evolution?.}

\com{Niemann, U.; Völzke, H.; Kühn, J.-P. Spiliopoulou, M. (2014), Learning and inspecting classification rules from longitudinal epidemiological data to identify predictive features on hepatic steatosis, Expert Systems with Applications 41 (11) , 5405-5415 .}

\com{Glaßer, S., Niemann, U., Preim, B., and Spiliopoulou, M. (2013). Can we Distinguish Between Benign and Ma- lignant Breast Tumors in DCE-MRI}

\com{Uli, can you write a short summary on these?}

\paragraph{Statistical Analysis}
\com{Regression Cube: A Technique for Multidimensional Visual Exploration and Interactive Pattern Finding}

\com{[Hierarchical Brushing of High-Dimensional Data Sets Using Quality Metrics]}

\com{Quality Metrics in High-Dimensional Data Visualization - An Overview and Systematization}

\paragraph{Visual Analysis of Heterogenous Data.}
Zhang et al. \cite{Zhang2012} provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set.
%
Their definition of a cohort differs from the understanding of the term in an epidemiological context by denoting every parameter-divided subject group as individual cohort.
%
Due to the short paper length, detail is missing on the data types and their algorithms of identifying similar subjects or whether they employ statistical measures.
%
We employ the idea of adding variables via drag and drop into a canvas area.

@@

\section{Implementation} \label{implementation}
\begin{figure}[htb]
 \centering
 \includegraphics[width=2.8in]{figures/placeholder}
 \caption{
 \com{Describe the Bootstrap/Angular/D3 Frontend, Node backend, R/OCPU Backend.}
 }
  \label{fig:Implementation}
\end{figure}
%%
Relying on web technologies paid off in previous collaboration with our clinical exerts, since it allows for:
\begin{itemize}
	\item easy exchange of software, without the need of installing anything other than a modern web browser,
	\item fast iterative adaption of method changes, without the need of installing updates,
	\item outsourcing computation-heavy tasks to server-clusters, which do not have to be in local proximity.
\end{itemize}
%%
The ongoing transition of open-science software into the web spawned numerous projects, making state-of-the-art algorithms available in this domain.
%%
\paragraph{Front-End.}
%%
The front end is created using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%%
\texttt{Angular.js}\footnote{Open Source; Maintained by Google, \href{https://www.angularjs.org/}{\texttt{angularjs.org}}} abstracts web application into models and views, allowing for a responsive way to combine \texttt{HTML} and \texttt{Javascript}.
%%
\texttt{Angular.js} is forcing developers to write modularized code, which makes the components easier expandable while keeping the code maintainable by including unit tests.
%%
The page layout is handled using \texttt{Twitter Bootstrap}\footnote{Open Source; Maintained by Twitter, \href{http://getbootstrap.com}{\texttt{getbootstrap.com}}}, which also provides a rich set of UI elements with proper stylings.
%%
The 2D mosaic plot is implemented using \texttt{Data driven Documents}(\texttt{D3.js}), which is very popular in information visualization using vector graphics \cite{D3}.
%%
It provides fast and easy methods for binding data to graphical elements.
%%
The 3D plot is created using the WebGL-based \texttt{Threejs}\footnote{Open Source; Originally developed by R. Cabello, \href{http://threejs.org}{\texttt{threejs.org}}} library.
%%
We experimented with different ways for achieving the cube representation, including volume rendering, cube primitives for each data point and shader-based solutions.
%%
Proper open source volume rendering methods are not available for the web, the effort of writing one from scratch was not worth the counter value.
%%
Creating a cube primitive for each data point resulted in non-interactive frame-rates for data sets larger than 30 features (creating 30$^3$ cube primitives).
%%
We use a shader-based solution by rendering the cube as a sprite based particle system, allowing to customize color and opacity of every data point.
%%
It also is the fastest solution that we tested.

@@

\paragraph{Access and Source.}
%%
A running instance of the \emph{Regression Cube} prototype can be found under \href{http://regressioncube.herokuapp.com/}{\texttt{regressioncube.herokuapp.com}}.
%%
The source for the prototype is freely available at \texttt{Github}\footnote{\texttt{R}-based back-end: \href{https://github.com/paulklemm/regression-cube-r-package}{\\\texttt{github.com/paulklemm/regression-cube-r-package}}}\footnote{Front-End and Nodejs Webserver: \href{https://github.com/paulklemm/regression-cube-prototype}{\texttt{\\github.com/paulklemm/regression-cube-prototype}}}.
%%
Instructions and code on how to create a setup running the \emph{Regression Cube} statistical backend through a \texttt{Ubuntu} server using \texttt{OpenCPU} are referenced in the repository.
%%
The front-end can be deployed using \href{https://www.heroku.com/}{\texttt{www.heroku.com}} by cloning the repository into a Heroku homepage. 
%%
%The front-end source as well as the \texttt{Nodejs} backend can be found here: .
%%
% \com{Use VAST'14 as Guideline}
% \com{Cloud computing}
% \com{Volume Rendering vs Particles}
% \com{Speed is merely a matter of available server machines due to parallelization process.}
% \com{Security enabled by hashing files.}
% \com{Bootstrap/Angular/D3 Frontend, Node backend, R/OCPU Backend (Fig.~\ref{fig:Implementation})}

@@

\section{Summary and Conclusion}
We presented a technique for knowledge discovery in cohort study data sets with user-defined target features.
%%
\emph{Regression Cube} are based on eponymous regression models, which allow to model domain knowledge about the data (e.g. influencing features on a disease) using formulas.
%%
Alternatively, \emph{Regression Cubes} can be calculated using default formulas to assess the explanatory power towards each feature in the data set.
%%
These two approaches allow both for a \emph{hypothesis-free} and \emph{hypothesis-based} explorative analysis.
%%
Correlation based feature selection reduces the amount of calculations using the target feature by focusing on the important regression models.
%%
The prototype described in this work was developed using state of the art web technologies, free services and open source libraries.
%%
All code associated to this project is also open source.
%%
Two case studies revealed that the presented approach is able to reproduce knowledge extracted using decision tree based data mining methods on a hepatic steatosis data set as well as producing new hypotheses by deriving insight into influencing factors on breast fat tissue in a explorative analysis.
%%

As future work, we want to introduce more regression models to the data set, which model different correlation types.
%%
We also want to extend the cube to time-dependent data by expanding the difference-cube approach.
%%

The method was very well received with our project partners, allowing them for the first time to retrieve a overview visualization of features, which does not highlights their pairwise correlations, but rather their explanatory power towards a target feature.
%%
The observed relationship are now subject of detailed statistical analyses.
%%
Our provided system not restricted to the epidemiological domain, using the provided instance, everybody can upload their own data or even set up their own \emph{Regression Cube} server cluster.
%%
We believe in the power of opening up knowledge discovery to allow a heterogenous group of domain experts to derive insight into their data and support the notion of open science.
%%
%%\paragraph{Future Work.}
%%\com{Implementing of regression analysis on the graphics card.}
%%
% \begin{small}
	%\acknowledgments{Omitted due to blind review}
\acknowledgments{SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no. 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania. Whole-body MR imaging was supported by a joint grant from Siemens Healthcare, Erlangen, Germany and the Federal State of Mecklenburg-Vorpommern. The University of Greifswald is a member of the ‘Centre of Knowledge Interchange’ program of the Siemens AG. This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics. We thank Marko Rak and Klaus Toennies for providing the image detection data.}
% \end{small}
\clearpage
\newpage
\bibliographystyle{abbrv}
%%use following if all content of bibtex file should be shown
%\nocite{*}
\bibliography{bibliography}
\end{document}
